---
layout: post
title: Air pollution in Karachi 
date: "30/04/2017"
tags: [Social Data Science, Air pollution, Data Visualization]
bigimg: 
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=4, echo=FALSE, warning=FALSE, message=FALSE, fig.align = 'center')
```

A friend of mine ([Max](http://maxkrueger.com/)) was discussing about his presentation on Smart Cities. While going through it, he mentioned a twitter account that posts about air pollution data about Karachi: [KarachiAir](https://twitter.com/KarachiAir). The tweets are machine generated and are posted every hour. It's a great step towards open data community and was a pleasent surprise for me to find it for Pakistani cities. 
Unfortunately, it is difficult to look at the big picture by looking at these tweets. Therefore, we thought about creating some visualizations using this data with some basic exploratory analysis. So we scraped the data from the twitter account using its [timeline API](https://dev.twitter.com/rest/reference/get/statuses/user_timeline); filtered it to get only quality measure tweets; parsed those tweets to get pm2.5 and air quality index and did some time series visualizations. 

## Exploration of Air pollution data in Karachi

```{r echo = FALSE}
library(xts)
library(broom)
library(dplyr)
library(ggplot2)
load("~/Personal/blogData/polKarachi.RData")
polDat$time = as.POSIXct(format(as.POSIXct(polDat$created,'PKT'), tz="Asia/Karachi"))
tsPol = xts(x = polDat$pm2.5Num,polDat$time)
PollutionTS = cbind(tsPol, polDat$airQNum)
names(PollutionTS) <- c("pm2.5","Air Quality Index")

dailyIndexs = apply.daily(PollutionTS, colMeans)
# index(PollutionTS) <- as.Date(index(PollutionTS))
tPollutionTS = tidy(PollutionTS)
# tPollutionTS %>% ggplot(aes(x=index,y=value, color=series)) + geom_line()
dailyPoll = tidy(dailyIndexs)
```

```{r echo=FALSE,results='asis',tidy=TRUE}
library(googleVis)
library(plyr)

filterPm2.5 <- function(PMvalue){
    if(PMvalue > 0 & PMvalue <= 30){
        "Good"}
    else if(PMvalue > 30 & PMvalue <= 60){
            "Satisfactory"
        }
    else if(PMvalue > 60 & PMvalue <= 90){
            "Moderately Polluted"
        }
    else "Severe"
}

filterAQI <- function(AQI){
    if(AQI > 0 & AQI <= 50.0){
        "Good"}
    else if(AQI > 50.0 & AQI <= 100.0){
            "Satisfactory"
        }
    else if(AQI > 100.0 & AQI <= 150){
            "Lightly Polluted"
        }
    else if(AQI > 150 & AQI <= 200){
            "Moderately Polluted"
        }
    else "Severe"
}


tPollutionTS$hazard = c(unlist(llply((tPollutionTS %>% dplyr::filter(series == "pm2.5") %>% dplyr::select(value))$value, filterPm2.5)),
                        unlist(llply((tPollutionTS %>% dplyr::filter(series == "Air.Quality.Index") %>%
                                      dplyr::select(value))$value, filterAQI)))


aPlot <-  gvisAnnotationChart(tPollutionTS, datevar="index",
                      numvar="value", idvar="series",
                      options=list(
                        width=600, height=350,
                        fill=10, displayExactValues=TRUE,
                        colors="['#0000ff','#00ff00']"),
                      chartid="AnnotationChart"
  )

aPlot$html$header <- NULL
aPlot$html$footer <- NULL
aPlot$html$jsFooter <- NULL
aPlot$html$caption <- NULL
print(aPlot)

```

It seems that the air quality index is increasing with the passage of time, whereas the pm2.5 metric is decreasing. If we pass our data through a moving average filter by using a one day window, this trend becomes more evident. 

```{r echo=FALSE,results='asis',tidy=TRUE}

dailyPoll$hazard = c(unlist(llply((dailyPoll %>% dplyr::filter(series == "pm2.5") %>% dplyr::select(value))$value, filterPm2.5)),
                                unlist(llply((dailyPoll %>% dplyr::filter(series == "Air.Quality.Index") %>%
                                      dplyr::select(value))$value, filterAQI)))
aPlot2 <-  gvisAnnotationChart(dailyPoll, datevar="index",
                      numvar="value", idvar="series",
                      options=list(
                        #width=600, height=350,
                        fill=10, displayExactValues=TRUE,
                        colors="['#0000ff','#00ff00']"),
                      chartid="AnnotationChart1"
                      )

aPlot2$html$header <- NULL
aPlot2$html$footer <- NULL
aPlot2$html$jsFooter <- NULL
aPlot2$html$caption <- NULL
print(aPlot2)

```

Finally, we look at the motion chart for this data, along with hazard levels based on __pm2.5 values__ and __air quality indexes__ as defined by [BlissAir](https://blissair.com/what-is-pm-2-5.htm).

```{r echo=FALSE,results='asis',tidy=TRUE}

dailyPoll$index = as.Date(dailyPoll$index)
dailyPollpm = dailyPoll %>% dplyr::filter(series == "pm2.5")
dailyPollaq = dailyPoll %>% dplyr::filter(series == "Air.Quality.Index")
dailyPollpm = dailyPollpm[!duplicated(dailyPollpm[,c('index')]),]
dailyPollaq = dailyPollaq[!duplicated(dailyPollaq[,c('index')]),]
dailyPoll2 = rbind.data.frame(dailyPollpm,dailyPollaq)

state2 = '
{"duration":{"multiplier":1,"timeUnit":"D"},
"yAxisOption":"2","showTrails":true,"playDuration":15000,
"sizeOption":"_UNISIZE","xAxisOption":"_TIME",
"xZoomedDataMin":1484092800000,"orderedByY":false,"xZoomedIn":false,
"time":"2017-01-11","xLambda":1,"orderedByX":false,
"yZoomedIn":false,"yZoomedDataMax":188.2,"yZoomedDataMin":10.65,
"nonSelectedAlpha":0.4,"uniColorForNonSelected":false,
"iconKeySettings":[{"trailStart":"2017-01-11","key":{"dim0":"Air.Quality.Index"}}],
"yLambda":1,"colorOption":"3","dimensions":{"iconDimensions":["dim0"]},"xZoomedDataMax":1493510400000,"iconType":"BUBBLE"}
'

Motion=gvisMotionChart(dailyPoll2, 
                       idvar="series", 
                       timevar="index",
                       options=list(state = state2))

Motion$html$header <- NULL
Motion$html$footer <- NULL
Motion$html$jsFooter <- NULL
Motion$html$caption <- NULL
print(Motion)

```
